---
title: "RNA-seq analysis in R"
subtitle: "Alignment and feature counting"
author: "Maria Doyle, Harriet Dashnow, Belinda Phipson"
date: "`r format(Sys.time(), '%d %B %Y')`"
bibliography: ref.bib
minutes: 60
---

# Introduction

Mapping reads to the genome is a very important task, and many different aligners are available, such as bowtie [@Langmead2012], topHat [@trapnell2009tophat], STAR [@Dobin2013] and Rsubread [@liao2013subread]. Rsubread is the only aligner that can run in R. Most mapping tasks require larger computers than an average laptop, so usually read mapping is done on a server in a Linux-like environment. However, for the purpose of this workshop, we will demonstrate how read alignment and counting can be performed using R. We will map a small subset of reads (1000) from each sample from our mouse lactation dataset [@Fu2015], and we will only be mapping to chromosome 1. This is so that everyone can have a go at alignment and counting using R.  

## Acknowledgements

This material was inspired by this [edgeR case study]((http://www.statsci.org/smyth/pubs/QLedgeRPreprint.pdf)) by [@Lun2016] and this [Rsubread case study](http://bioinf.wehi.edu.au/RNAseqCaseStudy/).

## Bioconductor/R Packages

Packages used:

* Rsubread

To install the package you can:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager"))
    install.packages("BiocManager")
BiocManager::install("Rsubread")
```

## Data files

* Mouse chromosome 1 Rsubread index files (~400MB).  
* 12 fastq.gz files each containing 1000 reads (~1MB)

Mouse mammary data (fastq files): [https://figshare.com/s/f5d63d8c265a05618137](https://figshare.com/s/f5d63d8c265a05618137)
You should download these files and place them in your `/data` directory. See below for how you can import from the links using R.

The raw reads were downloaded from SRA from the link given in [GEO](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE60450) for the dataset - ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP%2FSRP045%2FSRP045534. These files are in .sra format. The sra toolkit from NCBI was used to convert the .sra files to .fastq files using the fastq-dump command and the first 1000 reads in each file was extracted.

## Downloading genome files

We have provided the index files for chromosome 1 for the mouse genome build mm10 for this workshop in order to save time on building the index. However, full genome fasta files for a number of different genomes are available to download from the UCSC genome browser, see http://hgdownload.soe.ucsc.edu/downloads.html; from NCBI: http://www.ncbi.nlm.nih.gov/genome; or from ENSEMBL: http://asia.ensembl.org/info/data/ftp/index.html.

## Importing the data

First, let's load the Rsubread package into R.

```{r}
library(Rsubread)
```


```{r echo=FALSE}
# create data directory to download fastqs into for rendering this doc
dir.create("data") 
```

```{r message=FALSE}
# links to fastq files
fastq_urls <- c("https://ndownloader.figshare.com/files/5053552?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053585?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053582?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053579?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053588?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053555?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053558?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053561?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053564?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053567?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053570?private_link=f5d63d8c265a05618137",
                "https://ndownloader.figshare.com/files/5053573?private_link=f5d63d8c265a05618137")
          
# names to save files as        
fastq_names <- c("SRR1552444.fastq.gz", 
                 "SRR1552445.fastq.gz",
                 "SRR1552446.fastq.gz",
                 "SRR1552447.fastq.gz",
                 "SRR1552448.fastq.gz",
                 "SRR1552449.fastq.gz",
                 "SRR1552450.fastq.gz",
                 "SRR1552451.fastq.gz",
                 "SRR1552452.fastq.gz",
                 "SRR1552453.fastq.gz",
                 "SRR1552454.fastq.gz",
                 "SRR1552455.fastq.gz")

# download files into data directory
for(i in seq_along(fastq_urls )){
    download.file(fastq_urls[i], paste0("data/", fastq_names[i]), mode="wb")
}
```


```{r eval=FALSE}
# download mouse chr1 index files
# these files are relatively large (~400MB total) so it may take a few minutes, the default timeout for download.file is 60 secs so you may need to increase it. We'll increase to 100 secs.
options(timeout=100)

chr1index_urls <- c("https://ndownloader.figshare.com/files/5053597?private_link=f5d63d8c265a05618137",
                    "https://ndownloader.figshare.com/files/5053600?private_link=f5d63d8c265a05618137",
                    "https://ndownloader.figshare.com/files/5053606?private_link=f5d63d8c265a05618137",
                    "https://ndownloader.figshare.com/files/5053603?private_link=f5d63d8c265a05618137")

chr1index_names <- c("chr1_mm10.00.b.array", 
                     "chr1_mm10.00.b.tab",
                     "chr1_mm10.files",
                     "chr1_mm10.reads")

for(i in seq_along(chr1index_urls)){
    download.file(chr1index_urls[i], paste0("data/", chr1index_names[i]), mode="wb")
}
```

We put all the sequencing read data (.fastq.gz files) in the data directory.
Now we need to tell the Rsubread aligner which files to look at.
We can search for all .fastq.gz files in the data directory using the `list.files` command.
The pattern argument takes a regular expression.
In this case we are using the `$` to mean the end of the string, so we will only get files that end in ".fastq.gz"

```{r}
fastq.files <- list.files(path = "./data", pattern = ".fastq.gz$", full.names = TRUE)
fastq.files
```

## Alignment

### Build the index

Read sequences are stored in compressed (gzipped) FASTQ files. Before the differential expression analysis can proceed, these reads must be aligned to the mouse genome and counted into annotated genes. This can be achieved with functions in the Rsubread package.

The first step in performing the alignment is to build an index. In order to build an index you need to have the fasta file (.fa), which can be downloaded from the UCSC genome browser. Here we are building the index just for chromosome 1. This may take several minutes to run. Building the full index using the whole mouse genome usually takes about 30 minutes to an hr on a server. We won't be building the index in the workshop due to time constraints, we have provided the index files for you. The command below assumes the chr1 genome information for mm10 is stored in the "chr1.fa" file.

```{r,results="hide"}
# buildindex(basename="chr1_mm10",reference="chr1.fa")
```

The above command will generate the index files in the working directory. In this example, the prefix for the index files is chr1_mm10. You can see the additional files generated using the `dir` command, which will list every file in your current working directory.

```{r,results="hide"}
dir()
```

### Aligning reads to chromosome 1 of reference genome

Now that we have generated our index, we can align our reads using the `align` command. There are often numerous mapping parameters that we can specify, but usually the default mapping parameters for the `align` function are fine. If we had paired end data, we would specify the second read file/s using the `readfile2` argument. Our mouse data comprises 100bp single end reads.

We can specify the output files, or we can let `Rsubread` choose the output file names for us. The default output file name is the filename with ".subread.BAM" added at the end.

Now we can align our 12 fastq.gz files using the `align` command.

```{r, eval=FALSE}
align(index="data/chr1_mm10",readfile1=fastq.files)
```

This will align each of the 12 samples one after the other. As we're only using a subset of 1000 reads per sample, aligning should just take a minute or so for each sample. To run the full samples from this dataset would take several hours per sample. The BAM files are saved in the working directory.

To see how many parameters you can change try the `args` function:

```{r}
args(align)
```

In this example we have kept many of the default settings, which have been optimised to work well under a variety of situations. The default setting for `align` is that it only keeps reads that uniquely map to the reference genome. For testing differential expression of genes, this is what we want, as the reads are unambiguously assigned to one place in the genome, allowing for easier interpretation of the results. Understanding all the different parameters you can change involves doing a lot of reading about the aligner that you are using, and can take a lot of time to understand! Today we won't be going into the details of the parameters you can change, but you can get more information from looking at the help:

```{r, eval=FALSE}
?align
```

We can get a summary of the proportion of reads that mapped to the reference genome using the `propmapped` function.

```{r eval=FALSE}
bam.files <- list.files(path = "./data", pattern = ".BAM$", full.names = TRUE)
bam.files
```

```{r echo=FALSE}
# locate precreated bam files for rendering this doc
bam.files <- list.files(path = system.file("extdata/bams", package = "RNAseqRworkshop"), pattern = ".BAM$", full.names = TRUE)

# print names emulating what they should see when running the chunk above
printfiles <- basename(bam.files)
paste0("./data/", printfiles)
```

```{r}
props <- propmapped(files=bam.files)
props
```

<div id="challenge">
<i class="fas fa-pencil-alt"></i> **Challenge**

1. Try aligning the fastq files allowing multi-mapping reads (set `unique = FALSE`), and allowing for up to 6 "best" locations to be reported (`nBestLocations = 6`). Specify the output file names (bam.files.multi) by substituting ".fastq.gz" with ".multi.bam" so we don't overwrite our unique alignment bam files.
2. Look at the proportion of reads mapped and see if we get any more reads mapping by specifying a less stringent criteria.

</div>

## Quality control

We can have a look at the quality scores associated with each base that has been called by the sequencing machine using the `qualityScores` function in *Rsubread*.

Let's first extract quality scores for 100 reads for the file "SRR1552450.fastq.gz".

```{r}
# Extract quality scores
qs <- qualityScores(filename="data/SRR1552450.fastq.gz",nreads=100)
# Check dimension of qs
dim(qs)
# Check first few elements of qs with head
head(qs)
```

A quality score of 30 corresponds to a 1 in 1000 chance of an incorrect base call. (A quality score of 10 is a 1 in 10 chance of an incorrect base call.) To look at the overall distribution of quality scores across the 100 reads, we can look at a boxplot

```{r}
boxplot(qs)
```

<div id="challenge">
<i class="fas fa-pencil-alt"></i> **Challenge**

 1. Extract quality scores for SRR1552451.fastq.gz for 50 reads.
 2. Plot a boxplot of the quality scores for SRR1552451.fastq.gz.

</div>

## Counting

Now that we have figured out where each read comes from in the genome, we need to summarise the information across genes or exons. The alignment produces a set of BAM files, where each file contains the read alignments for each library. In the BAM file, there is a chromosomal location for every read that mapped uniquely. The mapped reads can be counted across mouse genes by using the `featureCounts` function. `featureCounts` contains built-in annotation for mouse (mm9, mm10) and human (hg19) genome assemblies (NCBI refseq annotation).

The code below uses the exon intervals defined in the NCBI refseq annotation of the mm10 genome. Reads that map to exons of genes are added together to obtain the count for each gene, with some care taken with reads that span exon-exon boundaries. `featureCounts` takes all the BAM files as input, and outputs an object which includes the count matrix, similar to the count matrix we have been working with on Day 1. Each sample is a separate column, each row is a gene.

```{r,results="hide"}
fc <- featureCounts(bam.files, annot.inbuilt="mm10")
```

```{r}
# See what slots are stored in fc
names(fc)
```

The statistics of the read mapping can be seen with fc$stats. This reports the numbers of unassigned reads and the reasons why they are not assigned (eg. ambiguity, multi-mapping, secondary alignment, mapping quality, fragment length, chimera, read duplicate, non-junction and so on), in addition to the number of successfully assigned reads for each library. (We know the real reason why the majority of the reads aren't mapping - they're not from chr 1!)

```{r}
## Take a look at the featurecounts stats
fc$stat
```

The counts for the samples are stored in fc$counts. Take a look at that.

```{r}
## Take a look at the dimensions to see the number of genes
dim(fc$counts)
## Take a look at the first 6 lines
head(fc$counts)
```

The row names of the fc$counts matrix represent the Entrez gene identifiers for each gene and the column names are the output filenames from calling the `align` function. The `annotation` slot shows the annotation information that `featureCounts` used to summarise reads over genes.

```{r}
head(fc$annotation)
```

<div id="challenge">
<i class="fas fa-pencil-alt"></i> **Challenge**

 1. Redo the counting over the exons, rather than the genes (specify `useMetaFeatures = FALSE`). Use the bam files generated doing alignment reporting only unique reads, and call the `featureCounts` object `fc.exon`. Check the dimension of the counts slot to see how much larger it is.
 2. Using your ".multi.bam" files, redo the counting over genes, allowing for multimapping reads (specify `countMultiMappingReads = TRUE`), calling the object `fc.multi`. Check the stats.

</div>

Notes

* If you are sequencing your own data, the sequencing facility will almost always provide fastq files.  
* For publicly available sequence data from GEO/SRA, the files are usually in the Sequence Read Archive
(SRA) format. Prior to read alignment, these files need to be converted into the
FASTQ format using the fastq-dump utility from the SRA Toolkit. See http:
//www.ncbi.nlm.nih.gov/books/NBK158900 for how to download and use the
SRA Toolkit.  
* By default, alignment is performed with `unique=TRUE`. If a read can be aligned to
two or more locations, *Rsubread* will attempt to select the best location using a
number of criteria. Only reads that have a unique best location are reported as
being aligned. Keeping this default is recommended, as it avoids spurious signal
from non-uniquely mapped reads derived from, e.g., repeat regions.  
* The Phred offset determines the encoding for the base-calling quality string in the
FASTQ file. For the Illumina 1.8 format onwards, this encoding is set at +33.
However, older formats may use a +64 encoding. Users should ensure that the
correct encoding is specified during alignment. If unsure, one can examine the
first several quality strings in the FASTQ file. A good rule of thumb is to check
whether lower-case letters are present (+64 encoding) or absent (+33).  
* `featureCounts` requires gene annotation specifying the genomic start and end
position of each exon of each gene. *Rsubread* contains built-in gene annotation
for mouse and human. For other species, users will need to read in a data frame
in GTF format to define the genes and exons. Users can also specify a custom annotation file in SAF format. See the Rsubread users guide for more information, or try `?featureCounts`, which has an example of what an SAF file should like like.

# Package versions used

```{r}
sessionInfo()
```

# References
